---
title: "Routing and Trajectories"
author: "Nathan Hoteling"
date: "`r format(Sys.time(), '%B %Y')`"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(tidyr)
library(stringr)
library(stringdist)
library(osmdata)
library(osrm)
library(ggplot2)
library(h3)
library(RcppRoll)
library(readr)

library(kableExtra)

library(sp)
library(sf)
library(rnaturalearth)        # to get geo shapes
library(rnaturalearthhires)   # same as above

library(spacetime)
library(STPtrajectories)

source("../R/utils.R")
```

```{r data, echo=FALSE, message=FALSE, warning=FALSE}
# us_states_cont <- ne_states(
#     country = 'united states of america',
#     returnclass = 'sf') %>%
#     filter(! name %in% c('Alaska', 'Hawaii'))
# 
# saveRDS(us_states_cont, "data/us_states_cont.rds")

us_states_cont <- readRDS("../data/us_states_cont.rds")
us <- sf::st_union(us_states_cont)

df.rte <- readRDS("../data/df_rte.rds")
```


<hr>
<br>

## Introduction

This write-up includes an investigation into trajectory mining with spatio-temporal data, primarily with techniques derived from [Miller's space-time prisms](https://cpb-us-w2.wpmucdn.com/u.osu.edu/dist/a/390/files/2017/03/Miller-time-geography-and-space-time-prism-1rp3u1q.pdf) (STPs) and implemented with the `STPtrajectories` [R package](https://github.com/markvregel/STPtrajectories).   


```{r, eval=FALSE, echo=FALSE}
## Note: install failed when I tried to build vignettes
devtools::install_github("markvregel/STPtrajectories",build_vignettes=FALSE)
```

<br>


## Data Simulation

A simulated dataset is generated from the travel route between Portland, OR and Portland, ME.  City coordinates are obtained from OpenStreetMap via `osmdata` and the route between cities from `osrm`.  Five hundred imaginary trips are simulated from this route by sampling from a Normal distribution with centroid value being the point coordinates.  The resulting point cloud is displayed below, with the original path depicted as a black line.


```{r, echo=FALSE, message=FALSE, warning=FALSE, eval=FALSE}

# Coords for  Portland, OR
pdx <- osmdata::getbb("Portland, OR", format_out="sf_polygon") 
pt_pdx <- pdx$multipolygon %>% sf::st_centroid()

# Coords for Portland, ME
ptm <- osmdata::getbb("Portland, ME", format_out="sf_polygon")
pt_ptm <- ptm$multipolygon %>% sf::st_centroid()

# Route from Portland to Portland
tstart <- as.POSIXct("2022-07-01 09:00")
avg_spd <- 50 # km/h
std_spd <- 10
df.rte <- osrm::osrmRoute(src=pt_pdx, dst=pt_ptm) %>%
  mutate(pt  = crd_to_points(lon,lat),
         dst = sf::st_distance(pt, lag(pt), by_element=TRUE),
         tm  = tstart)



for (i in 2:nrow(df.rte)) {
    t_old <- df.rte$tm[i-1]
    v <- rnorm(1, avg_spd, std_spd)
    x <- ((df.rte$dst[i]/1000) / v) %>% as.numeric() %>% round(digits=2)
    
    v1 <- (x-floor(x)) - 0.60
    v2 <- ifelse(v1 > 0, floor(x) + 1 + v1, x)
    HrMn = sub(".", ":", v2, fixed = TRUE) %>% hm()
    
    df.rte$tm[i] <- df.rte$tm[i-1] + HrMn
    }
  


saveRDS(df.rte, "../data/df_rte.rds")
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Simulate some data
epsg_code <- spaceheater::get_epsg(df.rte$pt)
df.rte1 <- df.rte %>% mutate(pt1 = pt %>% sf::st_transform(epsg_code))
xx <- sf::st_coordinates(df.rte1$pt1)
df.rte1$lon1 <- xx[,1]
df.rte1$lat1 <- xx[,2]

N <- 500
V_SD <- 75000 # sd in meters
set.seed(1234)  # So random numbers are repeatable
d <- lapply(1:nrow(df.rte), function(i) {
  lon <- rnorm(N,df.rte1$lon1[i],V_SD) #+ rbinom(N,3,0.2)
  lat <- rnorm(N,df.rte1$lat1[i],V_SD) #+ rbinom(N,2,0.2)
  idx_tr <- 1:N
  idx_pt <- i
  lonlat <- crd_to_points(lon,lat, CRS=epsg_code) %>% sf::st_transform(4326) %>% sf::st_coordinates()
  data.frame(lon=lonlat[,1], lat=lonlat[,2],idx_tr=idx_tr, idx_pt=idx_pt)
})
df.dat <- do.call(rbind,d) %>% 
  mutate(pt = crd_to_points(lon,lat),
         hex = h3::geo_to_h3(c(lat,lon), res=3)) %>%
  arrange(idx_tr,idx_pt) 
```


<br>


```{r, echo=FALSE, message=FALSE, warning=FALSE}
#
# Plot the route as a line and super-impose
# a sample of the simulated lines for illustrative purposes
#


ln <- make_line(sf::st_coordinates(df.rte$pt)) %>% sf::st_sfc(crs=4326)
df.tmp <- df.dat %>% 
  group_by(idx_tr) %>% 
  summarise(ln = make_line(sf::st_coordinates(pt)) %>% sf::st_sfc(crs=4326))
p1 <- df.dat %>% #df.tmp[sample(nrow(df.tmp),10),] %>%
  ggplot() +
  geom_sf(data=us, mapping=aes(geometry=geometry), fill="cornsilk", color="grey75", size=0.2) +
  geom_sf(aes(geometry=pt), alpha=0.3, color="grey65", size=0.5) +
  geom_sf(data=ln, aes(geometry=geometry)) +
  theme_void()

# p <- ggplot(data = us_states_cont) +
#   geom_sf(fill = "cornsilk", color = "grey80", size=0.2) +
#   theme_void()
#
# Plot a hex representation of all the simulated points
#
df.hex <- df.dat %>% 
  group_by(hex) %>% 
  count() %>%
  mutate(geometry = h3::h3_to_geo_boundary_sf(hex) %>% pull(geometry)) #%>%
p2 <- df.hex %>% 
  ggplot() +
  geom_sf(data=us, mapping=aes(geometry=geometry), fill="cornsilk", color="grey75", size=0.2) +
  geom_sf(aes(geometry=geometry, fill=n),color=NA) + 
  scale_fill_gradient(low="grey75",high="firebrick", guide="none") +
  theme_void()

```


```{r, echo=FALSE, message=FALSE, warning=FALSE, out.width="48%", fig.align="center"}
p1
#p2
```

<br>
<br>

## Monte Carlo from scratch

We can use the 500 imaginary trips to estimate the original route via monte carlo.  To do this, we aggregate the point data by geohash (via the `h3` package), and the model uses the _current_ geohash to predict the _next_ geohash.  The "model" in this case is really just a collection of geohash-pairs plus the number of times this pair was seen in the data.  To carry out the monte carlo process, we start at one end of the route and randomly sample from geohash pairs to determine the next grid location, until the process is completed (ie: we have arrived at the other end of the route, or some maximum number of iterations has been reached).  The procedure is repeated many times so that we have a collection of paths.


```{r, echo=FALSE, warning=FALSE, message=FALSE}
# Create hex pairs: hex to next-hex
model_stuff <- function(df) {
  df %>% rename(hex0=hex) %>%
    mutate(hex1 = lead(hex0, n=1), n=1)
}

# Build the correlation model
df.mod <- df.dat %>%
  dplyr::select(idx_tr, idx_pt, hex) %>%
  tidyr::nest(data=c(idx_pt,hex)) %>%
  mutate(data = purrr::map(data,model_stuff)) %>%
  unnest(cols=c(data)) %>%
  group_by(hex0,hex1) %>%
  summarise(w = sum(n)) %>%
  ungroup()
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}

# Pick the next hex based on the present hex
random_picker <- function(hex, df.model) {
  df <- df.model %>% filter(hex0==hex) #%>% mutate(z=w/sum(w))
  df$z <- df$w/sum(df$w)
  if (nrow(df) > 0) {
    hx <- sample(df$hex1, 1, prob=df$z)
  } else {hx <- NA}
}

# Get all the possible start points
start_points <- df.dat %>%
  filter(idx_pt == 1) %>%
  group_by(hex) %>%
  count() 
start_points$r <- start_points$n/sum(start_points$n)


# Monte Carlo function
my_monte_carlo <- function(start, max_iter, df.model) {
  d <- list()
  d[[1]] <- start
  for (i in 2:max_iter) {
    hex_new <- d[[i-1]]
    d[[i]] <- random_picker(hex_new, df.model)
    if (is.na(d[[i]])) {break}
  }
  M <- length(d)
  data.frame(idx=1:M, hex=unlist(d)) %>%
    filter(!is.na(hex)) %>%
    mutate(geometry = h3::h3_to_geo_boundary_sf(hex) %>% pull(geometry),
           pt = sf::st_centroid(geometry))
}
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Get a random start point
fstart <- sample(start_points$hex, 1, prob=start_points$r) 
df.mte <- my_monte_carlo(fstart, max_iter=30, df.mod)
ln2 <- df.mte %>% pull(pt) %>% sf::st_coordinates() %>% make_line() %>% sf::st_sfc(crs=4326)
```


```{r, echo=FALSE, out.width="40%", fig.align="center"}
px <- ggplot() +
  geom_sf(data=us, mapping=aes(geometry=geometry), fill="cornsilk", color="grey75", size=0.2) +
  geom_sf(data=ln, aes(geometry=geometry)) +
  geom_sf(data=ln2, aes(geometry=geometry), color="firebrick", alpha=0.5, size=0.5) +
  theme_void() 
```



```{r, echo=FALSE, message=FALSE, warning=FALSE}
N <- 10
fstart <- sample(start_points$hex, N, prob=start_points$r) 
dd <- lapply(1:N, function(i) {
  my_monte_carlo(fstart[i], max_iter=30, df.mod) %>% mutate(nn = i)
})
df.mte2 <- do.call(rbind,dd) %>%
  group_by(nn) %>%
  summarise(ln = sf::st_coordinates(pt) %>% make_line() %>% sf::st_sfc(crs=4326))
```


```{r, echo=FALSE, out.width="48%", fig.align="center"}
py <- px + geom_sf(data=df.mte2, aes(geometry=ln), color="firebrick", alpha=0.3, size=0.5)
py
```

<br>

### Next-Hex Probabilities

More interesting than the gross monte carlo results described above is a determination of the "next-hex probabilities".  In other words, given some current position, what does the probability distribution look like for each of the surrounding positions?  This is fairly straightforward to implement with the model developed in the section above.  In the example below, the current-hex is depicted in black, and the next-hex probabilities are represented by fill color in the surrounding geohashes, 


```{r, echo=FALSE, message=FALSE, warning=FALSE}
set.seed(1234)
fstart <- sample(start_points$hex, 1, prob=start_points$r) 
dfx <- df.mod %>% filter(hex0==fstart) 
dfx$z <- dfx$w/sum(dfx$w)

df.prb <- dfx %>%
  mutate(geometry = h3::h3_to_geo_boundary_sf(hex1) %>% pull(geometry))
```


```{r, echo=FALSE, message=FALSE, warning=FALSE, out.width="48%", fig.align="center"}
hex0 <- df.prb$hex0 %>% unique()
p.prb <- ggplot() +
  geom_sf(data=us, mapping=aes(geometry=geometry), fill="cornsilk", color="grey75", size=0.2) +
  geom_sf(data=df.hex, aes(geometry=geometry),fill="grey95",color=NA) +
  geom_sf(data=df.prb, aes(geometry=geometry, fill=z), color="grey99",size=0.1) +
  geom_sf(data=df.hex %>% filter(hex==hex0), aes(geometry=geometry), fill="grey10") +
  scale_fill_gradient(low="grey70",high="firebrick", guide="none") +
  theme_void()

p.prb
```


## Miller's Space-Time Prisms

> The space-time prism is the
> envelope of all possible space–time paths between
> known locations and times.

The R package [STPtrajectories](https://github.com/markvregel/STPtrajectories) includes an implementation of [Miller's space time prism](https://cpb-us-w2.wpmucdn.com/u.osu.edu/dist/a/390/files/2017/03/Miller-time-geography-and-space-time-prism-1rp3u1q.pdf) as well as some related features like a random trajectory generator, potential path area, and the alibi query.  Using the package [vignette](http://htmlpreview.github.io/?https://raw.githubusercontent.com/markvregel/STPtrajectories/master/vignettes/STP_Tracks.html) as a guide, we illustrate each of these features here within the context of the example above.


```{r, echo=FALSE, message=FALSE, warning=FALSE}
## Create a spacetime object from the route data

# note: STPtrajectories uses sp and not sf
# note: need to project coordinates for Track stuff to work below

sp_rte <- sf::as_Spatial(df.rte$pt %>% sf::st_transform(epsg_code))
stidf_rte <- STIDF(sp_rte, df.rte$tm, df.rte)

# Create a track object
# max speed is 25 m/s (~56mph)
track_rte <- STPtrajectories::STP_Track(Track(stidf_rte), 25)
```


### Potential Path Area

The potential path area (PPA) encompasses the region that can feasibly have been traversed between observed points, based on some maximum speed.  It is the spatial footprint of the space-time prism. Given a maximum average speed of around 25 m/s (~55 mph), the PPA for our path between Portlands is given in the chart below.   Notice that this looks quite different from the results presented above from the monte carlo simulation.  There is a good reason for this: the regions depicted operate on the assumption that data points represent observables and the regions represent how far an individual may have traveled between observed points, based on the maximum average speed between these points.  In contrast, the monte carlo results depicted above incorporate uncertainty on the observed positions, but ignore any variations between points.  In fact, these methods are complementary, in the sense that one might consider generating PPAs from the monte carlo paths, and then combining the results together into an overall path probability distribution.


```{r, echo=FALSE, message=FALSE, warning=FALSE, out.width="48%", fig.align="center"}
PPA_track_rte <- PPA(track_rte) %>% sf::st_as_sf() %>% sf::st_transform(4326)

p.ppa <- ggplot() + 
  geom_sf(data=us, mapping=aes(geometry=geometry), fill="cornsilk", color="grey75", size=0.2) +
  geom_sf(data=ln, aes(geometry=geometry)) +
  geom_sf(data=PPA_track_rte, mapping=aes(geometry=geometry), alpha=0.5, color="grey75") + theme_void()

p.ppa
```


### Random trajectory generator

In the area chart above the region around the Great Lakes spans a large area because the "observed" points are spread further from one another.  This means that, at 55 mph, an individual can travel further afield between observed points.  Ignoring the presence or absence of roads, one can get an idea of the potential paths within this area with the random trajectory generator (RTG).  The RTG uses a _directed random walk_ (ie: a random walk with directional bias) to generate a path between anchor points.  The chart below at left shows a single random trajectory generated to fill gaps between the "known" points.  We use this same process to simulate 500 hypothetical paths around the Great Lakes region, as displayed in the chart at right below.  Here, the paths are rendered as dots so that the result is effectively a dot-density map depicting the likelihood that an individual traveled through a given location.  The result, according to Miller, is a bivariate multinomial distribution centered on each of the anchor points.  Note that the density is noticeably smaller further from the markers, which represent observed locations.


```{r, echo=FALSE, message=FALSE, warning=FALSE, echo=FALSE}
df.rte2 <- df.rte[18:22,]
sp_rte2 <- sf::as_Spatial(df.rte2$pt %>% sf::st_transform(epsg_code))
stidf_rte2 <- STIDF(sp_rte2, df.rte2$tm, df.rte)
track_rte2 <- STPtrajectories::STP_Track(Track(stidf_rte2), 25)
```

```{r, message=FALSE, warning=FALSE, echo=FALSE, eval=FALSE}
## Run this once and save because it takes awhile
d <- lapply(1:500, function(i){
  STP_track_rte2 <- RTG(track_rte2, n_points = 10, iter=10)
  df <- sf::st_as_sf(STP_track_rte2@sp) %>% sf::st_transform(4326)
  #xy <- sf::st_coordinates(df$geometry)
  #xy_ln <- make_line(xy) 
})
df.rtg <- do.call(rbind,d)

d <- lapply(1:500, function(i){
  STP_track_rte2 <- RTG(track_rte2, n_points = 10, iter=10)
  df <- sf::st_as_sf(STP_track_rte2@sp) %>% sf::st_transform(4326)
  xy <- sf::st_coordinates(df$geometry)
  xy_ln <- make_line(xy) 
})
df.ln <- data.frame(ln = sf::st_as_sfc(d, crs=4326))

saveRDS(df.rtg, "../data/df_rtg_pt.rds")
saveRDS(df.ln,  "../data/df_rtg_ln.rds")
```


```{r, echo=FALSE, message=FALSE, warning=FALSE, out.width="48%", fig.show="hold"}
df.rtg1 <- readRDS("../data/df_rtg_pt.rds")
df.rtg2 <- readRDS("../data/df_rtg_ln.rds")
p.tst1 <- p.ppa + 
  geom_sf(data=df.rtg1, mapping=aes(geometry=geometry), color="firebrick", size=0.1, alpha=0.25) +
  geom_sf(data=df.rte2, mapping=aes(geometry=pt), color="grey25")
p.tst2 <- p.ppa + 
  geom_sf(data=df.rtg2[100,], mapping=aes(geometry=geometry), color="firebrick", size=0.3) +
  geom_sf(data=df.rte2, mapping=aes(geometry=pt), color="grey25")

p.tst2
p.tst1
```

### The Alibi Query

```{r, echo=FALSE, message=FALSE, warning=FALSE, eval=FALSE}
## NOT RUN ##
#  this chunk generates the route and saves a file to be read later

# Coords for  Fargo, ND
mnp <- osmdata::getbb("Fargo, ND", format_out="sf_polygon") 
pt_mnp <- mnp$geometry %>% sf::st_centroid() %>% sf::st_as_sf()

# Coords for Miami, FL
mia <- osmdata::getbb("Miami, FL", format_out="sf_polygon")
pt_mia <- mia$geometry %>% sf::st_centroid() %>% sf::st_as_sf()

# Route from Portland to Portland via osrm
tstart <- as.POSIXct("2022-07-01 12:00")
avg_spd <- 30 # km/h
std_spd <- 10
df.rte10 <- osrm::osrmRoute(src=pt_mnp, dst=pt_mia) %>%
  mutate(pt  = crd_to_points(lon,lat),
         dst = sf::st_distance(pt, lag(pt), by_element=TRUE),
         tm  = tstart)

## Simulate timing
for (i in 2:nrow(df.rte10)) {
    t_old <- df.rte10$tm[i-1]
    v <- rnorm(1, avg_spd, std_spd)
    x <- ((df.rte10$dst[i]/1000) / v) %>% as.numeric() %>% round(digits=2)
    
    v1 <- (x-floor(x)) - 0.60
    v2 <- ifelse(v1 > 0, floor(x) + 1 + v1, x)
    HrMn = sub(".", ":", v2, fixed = TRUE) %>% hm()
    
    
    df.rte10$tm[i] <- df.rte10$tm[i-1] + HrMn
    }

## save file  
saveRDS(df.rte10, "../data/df_rte10.rds")
```


```{r, echo=FALSE, message=FALSE, warning=FALSE}
## Read the file and convert to a Potential Path Area
df.rte10 <- readRDS("../data/df_rte10.rds")
sp_rte10 <- sf::as_Spatial(df.rte10$pt %>% sf::st_transform(epsg_code))
stidf_rte10 <- STIDF(sp_rte10, df.rte10$tm, df.rte10)
track_rte10 <- STPtrajectories::STP_Track(Track(stidf_rte10), 20)
PPA_track_rte10 <- PPA(track_rte10) %>% sf::st_as_sf() %>% sf::st_transform(4326)

# df.rte2 <- df.rte[18:22,]
# sp_rte2 <- sf::as_Spatial(df.rte2$pt %>% sf::st_transform(epsg_code))
# stidf_rte2 <- STIDF(sp_rte2, df.rte2$tm, df.rte2)
# track_rte2 <- STPtrajectories::STP_Track(Track(stidf_rte2), 25)


ln10 <- make_line(sf::st_coordinates(df.rte10$pt)) %>% sf::st_sfc(crs=4326)
# df.tmp <- df.dat %>% 
#   group_by(idx_tr) %>% 
#   summarise(ln = make_line(sf::st_coordinates(pt)) %>% sf::st_sfc(crs=4326))
p.path <- #df.dat %>% #df.tmp[sample(nrow(df.tmp),10),] %>%
  ggplot() +
  geom_sf(data=us, mapping=aes(geometry=geometry), fill="cornsilk", color="grey75", size=0.2) +
  #geom_sf(aes(geometry=pt), alpha=0.3, color="grey65", size=0.5) +
  geom_sf(data=ln, aes(geometry=geometry)) +
  geom_sf(data=PPA_track_rte, mapping=aes(geometry=geometry), alpha=0.5, color="grey75") +
  geom_sf(data=ln10, aes(geometry=geometry)) +
  geom_sf(data=PPA_track_rte10, mapping=aes(geometry=geometry), alpha=0.5, color="grey75", fill="grey75") + 
  theme_void()
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
## This takes a long time to run, so run once and save a file
d.mtg <- alibi_query(track_rte, track_rte10, stop_if_true=FALSE) ##, stop_if_true=FALSE, return_PIA=TRUE)

## The function returns a weird object so we need to extract info into a useful format
# d <- lapply(1:length(mtg), function(i){
#   if (!is.null(mtg[[i]]$PIA)) {
#     data.frame(geometry = mtg[[i]]$PIA %>% sf::st_as_sfc(),
#                trk11    = mtg[[i]]$STP_track1[1],
#                trk12    = mtg[[i]]$STP_track1[2],
#                trk21    = mtg[[i]]$STP_track2[1],
#                trk22    = mtg[[i]]$STP_track2[2],
#                tm1      = mtg[[i]]$meeting_time[1],
#                tm2      = mtg[[i]]$meeting_time[2])
#   }
# })
# df.mtg <- do.call(rbind, d)

## Save to a file
#saveRDS(df.mtg, "data/df_mtg.rds")

d <- lapply(1:length(d.mtg), function(i){
  a <- d.mtg[[i]]
  data.frame(track1 = a$STP_track1,
             track2 = a$STP_track2)
})
df.mtg <- do.call(rbind,d)

## There is probably a better way to do this but...
trk1 <- unique(df.mtg$track1) %>% sort()
sp_rte_x <- sf::as_Spatial(df.rte$pt[trk1] %>% sf::st_transform(epsg_code))
stidf_rte_x <- STIDF(sp_rte_x, df.rte$tm[trk1], df.rte[trk1,])
track_rte_x <- STPtrajectories::STP_Track(Track(stidf_rte_x), 25)
PPA_track_rte_x <- PPA(track_rte_x) %>% sf::st_as_sf() %>% sf::st_transform(4326)

trk2 <- unique(df.mtg$track2) %>% sort()
sp_rte_x2 <- sf::as_Spatial(df.rte10$pt[trk2] %>% sf::st_transform(epsg_code))
stidf_rte_x2 <- STIDF(sp_rte_x2, df.rte10$tm[trk2], df.rte10[trk2,])
track_rte_x2 <- STPtrajectories::STP_Track(Track(stidf_rte_x2), 20)
PPA_track_rte_x2 <- PPA(track_rte_x2) %>% sf::st_as_sf() %>% sf::st_transform(4326)

t_min <- max(c(min(df.rte$tm[trk1]), min(df.rte10$tm[trk2])))
t_max <- min(c(max(df.rte$tm[trk1]), max(df.rte10$tm[trk2])))
```

The alibi query uses the spatial intersection of two trajectories to determine whether they may have been at the same place at the same time, based on the mathematical approach described [here](https://www.tandfonline.com/doi/abs/10.1080/13658810902967397).  In the present case, we create a new route from Fargo, ND to Miami, FL, starting at noon on 01-JUL and traveling at an average speed of 30 km/h.  Given the two paths, the alibi query helps us determine a time period when the two trajectories may have intersected.  The chart below shows PPAs associated with the time periods of interest, with the darker shade representing the potential area of overlap.  In other words given what we know about the two paths, if these individuals met up at some point they likely did so within the dark red region some time between `r t_min` and `r t_max`.

```{r, echo=FALSE, message=FALSE, warning=FALSE, out.width="48%", fig.align="center"}
# df.mtg <- readr::read_rds("data/df_mtg.rds") %>%
#   mutate(geometry = sf::st_transform(geometry, 4326))

p.mtg <- ggplot() +
  geom_sf(data=us, mapping=aes(geometry=geometry), fill="cornsilk", color="grey75", size=0.2) +
  geom_sf(data=ln, aes(geometry=geometry)) +
  geom_sf(data=ln10, aes(geometry=geometry)) +
  geom_sf(data=PPA_track_rte_x, mapping=aes(geometry=geometry), fill="firebrick",alpha=0.25, color=NA) + 
  geom_sf(data=PPA_track_rte_x2, mapping=aes(geometry=geometry), fill="firebrick", alpha=0.25, color=NA) +
  theme_void()

p.mtg
```


### Potential Stay Time

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Coords for  Fargo, ND
slc <- osmdata::getbb("Salt Lake City, UT", format_out="sf_polygon") 
pt_slc <- slc$geometry %>% sf::st_centroid() %>% sf::st_as_sf()
buf_slc <- pt_slc$x[1] %>% sf::st_transform(epsg_code) %>% sf::st_buffer(50000) %>% sf::as_Spatial()

d.stay <- STPtrajectories::potential_stay(track_rte, buf_slc)
nm <- names(d.stay)
d <- lapply(1:length(d.stay), function(i){
  data.frame(nm  = nm[i],
             idx = readr::parse_number(nm[i]) %>% as.integer(),
             tm1 = d.stay[[i]][1],
             tm2 = d.stay[[i]][2])
})
df.stay <- do.call(rbind,d)


## There is probably a better way to do this but...
trkz <- unique(df.stay$idx) %>% sort()
sp_rte_z <- sf::as_Spatial(df.rte$pt[trkz] %>% sf::st_transform(epsg_code))
stidf_rte_z <- STIDF(sp_rte_z, df.rte$tm[trkz], df.rte[trkz,])
track_rte_z <- STPtrajectories::STP_Track(Track(stidf_rte_z), 25)
PPA_track_rte_z <- PPA(track_rte_z) %>% sf::st_as_sf() %>% sf::st_transform(4326)

t_min <- min(df.rte$tm[trkz])
t_max <- max(df.rte$tm[trkz])
```

With the information gathered above, we can determine a time interval in which a given trajectory intersects with a particular location.  For example, I'm curious when our original trajectory might pass within 50 km of Salt Lake City.  The chart below shows a 50-km buffer around Salt Lake City, with the PPAs associated with the time period that the individual may have been within this buffer.  The time period is `r t_min` to `r t_max`.

```{r, echo=FALSE, message=FALSE, warning=FALSE, out.width="48%", fig.align="center"}
slc_buffer <- buf_slc %>% sf::st_as_sf() %>%
  mutate(geometry = sf::st_transform(geometry, 4326))

p.stay <- ggplot() +
  geom_sf(data=us, mapping=aes(geometry=geometry), fill="cornsilk", color="grey75", size=0.2) +
  geom_sf(data=ln, aes(geometry=geometry)) +
  geom_sf(data=PPA_track_rte_z, mapping=aes(geometry=geometry), fill="firebrick",alpha=0.25, color=NA) + 
  geom_sf(data=slc_buffer, mapping=aes(geometry=geometry), fill="grey65", alpha=0.25, color="grey25") +
  theme_void()

p.stay
```


### Adding uncertainty to space and time

It is straightforward to recreate the Potential Path Area (PPA) above with uncertainty added to the time and position.  For the chart below, we add a position uncertainty of 10 km and a time uncertainty of 1 hour.  We also add an "activity time" of 20 minutes at each location. The result, as seen in the chart below, is that the PPA is porportionally larger. 

```{r, echo=FALSE, message=FALSE, warning=FALSE, out.width="48%", fig.align="center"}
track_rte100 <- STPtrajectories::STP_Track(track = Track(stidf_rte), 
                                           vmax  = 25, 
                                           activity_time = 20,
                                           location_uncertainty = 10000,
                                           time_uncertainty = 60)
PPA_track_rte100 <- PPA(track_rte100) %>% sf::st_as_sf() %>% sf::st_transform(4326)

p.trk100 <- ggplot() +
  geom_sf(data=us, mapping=aes(geometry=geometry), fill="cornsilk", color="grey75", size=0.2) +
  geom_sf(data=ln, aes(geometry=geometry)) +
  geom_sf(data=PPA_track_rte, mapping=aes(geometry=geometry), 
          fill="grey75",alpha=0.5, color=NA) + 
  geom_sf(data=PPA_track_rte100, mapping=aes(geometry=geometry), 
          fill="firebrick", alpha=0.25, color="grey25", size=0.2) +
  theme_void()

p.trk100
```


## Conclusions

Several useful and interesting capabilities are made available with Miller's space-time prisms, as implemented in the `STPtrajectories` R package.  

One potential drawback of the package is that is uses objects built on the older geospatial infrastructure from `sp` instead of the newer `sf`.  It may be possible to update this, however since many of the package-defined objects are built from other objects defined in the `spacetime` package, this could be more challenging that it appears to be on the surface.

There are several other approaches to trajectory mining in the literature that may be worth investigating, especially if there are already implementations available.  See _Other Interesting References_ in the References section below for more information.  Additionally, there are variations of the present methodology that account for road networks and other constraints, which may be worth exploring.

<br>

## References

<br>

<b>Primary References for this work</b>

The `STPtrajectories` R package:
[GitHub page](https://github.com/markvregel/STPtrajectories), 
[documentation](https://github.com/markvregel/STPtrajectories/blob/master/STPtrajectories.pdf), 
[vignette](http://htmlpreview.github.io/?https://raw.githubusercontent.com/markvregel/STPtrajectories/master/vignettes/STP_Tracks.html)  

[Time geography and space time prism](https://cpb-us-w2.wpmucdn.com/u.osu.edu/dist/a/390/files/2017/03/Miller-time-geography-and-space-time-prism-1rp3u1q.pdf), Miller, H. J. (2017)  

[What About People in Regional Science](http://courses.washington.edu/cee500/What%20about%20people%20in%20regional%20science.pdf), Lecture transcript from Torsten Hagerstrand (1970) ~ origin of time geography  

Analytic Solution for the alibi query: 
[official](https://www.tandfonline.com/doi/abs/10.1080/13658810902967397), 
[semantic scholar (for pdf)](https://www.semanticscholar.org/paper/An-analytic-solution-to-the-alibi-query-in-the-for-Kuijpers-Grimson/eecd8301548a50f15e100e0cdd7f132e5d17017d)  

<br>

<b>Other Interesting References</b>

[Spatio-temporal footprint change discovery: ](https://www-users.cse.umn.edu/~shekhar/talk/2014/changeFtprtDtcn.pdf)

[Exploring Movement: Similarity Analysis of Moving Objects](https://www.researchgate.net/publication/220351277_Exploring_movement_-_similarity_analysis_of_moving_objects) (2009)  

[Density-based co-location pattern discovery](https://www.researchgate.net/publication/221589564_Density_based_co-location_pattern_discovery) (2008)

[A neighborhood graph based approach to regional co-location pattern discovery](https://www.researchgate.net/publication/221589509_A_neighborhood_graph_based_approach_to_regional_co-location_pattern_discovery_A_summary_of_results) (2011)

[Discovering spatial co-location patterns: A summary of results](https://www-users.cse.umn.edu/~shekhar/research/co-location.pdf), Shekhar & Huang  

[Shashi Shekhar research page](https://www-users.cse.umn.edu/~shekhar/)  

[Nice R Geo lecture materials](http://eda.seas.gwu.edu/2021-Spring/c12-maps.html)  

TODO: 3D plots via [rayshader](https://www.tylermw.com/3d-ggplots-with-rayshader/)  

<br>
<br>
<hr>
<br>
<br>
<br>









